Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-07-04 write-ups/FSE/paper.tex
some work on the paper

\newcommand{\ADComment}[1]{\textcolor{green}{AD: #1}}
%\newcommand{\TSComment}[1]{\textcolor{blue}{TS: #1}}
\newcommand{\HEComment}[1]{\textcolor{orange}{HE: #1}}
%\newcommand{\COULDCUT}[1]{\textcolor{blue}{#1}}
\newcommand{\TSComment}[1]{\textcolor{purple}{TS: #1}}

@@

% lstlisting macros
\lstset{
  language=C, % choose the language of the code
  showspaces=false, % show spaces adding particular underscores
  showstringspaces=false, % underline spaces within strings
  showtabs=false, % show tabs within strings adding particular underscores
  tabsize=8, % sets default tabsize to 2 spaces
  captionpos=b, % sets the caption-position to bottom
  mathescape=true, % activates special behaviour of the dollar sign
  basicstyle=\footnotesize\tt,
  columns=fullfelxible,
  xleftmargin=2em,
  %commentstyle=\rmfamily\itshape,
  morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then,transmit},
  escapeinside={//(*@}{@*)},
  numbers=left
}

@@

@@

@@

%\authorinfo{Tyler Sorensen}
%{\makebox{Imperial College London, UK} \\
%}
%{t.sorensen15@imperial.ac.uk}

@@

%\newcommand{\myparagraph}[1]{\paragraph{#1.}}
\newcommand{\myparagraph}[1]{\vspace{1mm}\noindent\emph{\textbf{#1.}}}
%{\noindent{\bf\emph{#1}}\hspace{1em}}

@@

\myparagraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  There is growing interest in accelerating
such algorithms on GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.

@@

%\vspace{-1mm}
\myparagraph{A degree of fairness: occupancy-bound execution} The current GPU programming
models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding
scheduling of workgroups, and current GPU schedulers are unfair in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.\footnote{In practice,
  depending on the kernel, multiple workgroups might map to the same
  compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle more workgroups being
launched than there are compute units is via an \emph{occupancy-bound}
execution
model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16} where,
once a workgroup has commenced execution on a compute unit (it has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it finishes execution.
This is simple model is easy to implement
efficiently, and experiments suggest that
it is widely employed by today's
GPUs~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{overview.pdf}
\caption{Cooperative kernels can flexibly resize to let other tasks
(e.g. graphics) run concurrently}
\label{fig:overview}
\vspace{-2mm}
\end{figure}

@@

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that GPU vendors (1) will
not commit to the occupancy-bound execution model they currently
implement, for the above reasons, yet (2) will not guarantee fair
scheduling due to the runtime cost of preempting workgroups in a fair
manner.  Vendors instead wish to retain the essence of the simple
occupancy-bound model, supporting preemption only in key special
cases.

@@

A kernel
that requires fair scheduling is identified as \emph{cooperative}, and written using two additional
language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.
%
Where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates that the workgroup
would ideally continue executing, but that the
scheduler may preempt the workgroup; the cooperative kernel
must be prepared to deal with either scenario.
%
Where the cooperative kernel could use additional resources, a workgroup can execute
$\offerfork$ to indicates that the
kernel is prepared to proceed with the existing set of workgroups, but
is able to benefit from one or more additional workgroups
commencing execution directly after the $\offerfork$ program point.

@@

@@

In summary, our main contributions are:
\emph{cooperative kernels}, an extended GPU programming model that supports the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}); a \emph{prototype implementation} of cooperative
  kernels on top OpenCL 2.0
  (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms \cutthree{across three GPUs} (\mysec\ref{sec:experiments}), including a best-effort comparison with the efficiency afforded by hardware-supported preemption available on Nvidia GPUs.

@@

@@

\begin{lstlisting}
(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * queues) {
(*@\label{line:wksteal:getgroupid}@*)  int queue_id = get_group_id();
(*@\label{line:wksteal:mainloop}@*)  while (more_work(queues)) {
(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
    if (t)
(*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
  }
}
\end{lstlisting}
\vspace{-2mm}
\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
%\vspace{-6mm}
\end{figure}

@@

\myfiglong\ref{fig:workstealing} illustrates a work
stealing kernel. Each thread receives a pointer to the
task queues, in global memory, initialized by the
host to contain the initial tasks. A thread uses its workgroup id 
(line~\ref{line:wksteal:getgroupid}) as a queue id to access the relevant task queue. The
$\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
pops a task from the workgroup's queue or, if the queue is empty,
tries to steal a task from other queues.
%
% is called by all threads of a workgroup.  Inside
% $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
% the \emph{master thread}, tries to pop a
% task from the workgroup's queue, or to steal a task from a different
% queue. The master thread uses local memory inside
% $\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
% all threads of its workgroup.
%
If a task was obtained, then the workgroup processes it
(line~\ref{line:wksteal:processtask}), which may lead to new tasks
being created and pushed to the workgroup's queue.

Although not depicted here, concurrent accesses to queues inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex per queue, implemented using atomic compare and swap operations
on global memory.  The kernel presents two opportunities
for spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task.  The algorithm thus depends
on a fair scheduler, which is not guaranteed by current GPU programming models.

@@

\begin{lstlisting}
kernel graph_app(global graph * g,
                 global nodes * n0, global nodes * n1) {
  int level = 0;
  global nodes * in_nodes = n0;
  global nodes * out_nodes = n1;
  int tid = get_global_id();
  int stride = get_global_size();
(*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
    for (int i = tid; i < nodes.size; i += stride)
      process_node(g, in_nodes[i], out_nodes, level);
(*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
(*@\label{line:graph:gb1}@*)    global_barrier();
(*@\label{line:graph:reset}@*)    reset(out_nodes);
    level++;
(*@\label{line:graph:gb2}@*)    global_barrier();
  }
}
\end{lstlisting}
\vspace{-2mm}
\caption{An OpenCL graph traversal algorithm}\label{fig:graphsearch}
%\vspace{-2mm}
\end{figure}

@@

The global barrier used in this application is not provided as a GPU
primitive, though previous works have shown that such a global
barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16},
based on CPU barrier designs~\cite[ch. 17]{HS08}.
 These barriers employ spinning to ensure threads wait at the barrier until all
threads have arrived, which requires fair
scheduling between workgroups.% for the barrier to operate correctly.

The mutexes and barriers used by these two examples appear to run
reliably on current GPUs for kernels that are executed with no more
workgroups than there are compute units.  This is due to the fairness
of the occupancy-bound execution model that current GPUs have been
shown, experimentally, to provide.  But, as discussed in
\mysec\ref{sec:intro}, this model is not endorsed
language standards or vendor implementations, and is unlikely to be respected in the future.
%
In \mysec\ref{sec:programmingguidelines} we show how the work stealing
and graph traversal examples of \myfigs\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
updated to use our cooperative kernels programming model to resolve
the scheduling issue.

@@

%\vspace{-1mm}
\myparagraph{Active workroups}
%
If the host application launches a cooperative kernel requesting $N$
workgroups, this indicates that the kernel should be executed with a
maximum of $N$ workgroups, and that as many workgroups as possible, up
to this limit, are desired.  However, the scheduler may initially
schedule fewer than $N$ workgroups, and as explained below the number
of workgroups that execute the cooperative kernel can change during
the lifetime of the kernel.

@@

\begin{lstlisting}
kernel graph_app(global graph *g,
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < nodes.size; i += stride)
      process_node(g, in_nodes[i], out_nodes, level);
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
\vspace{-4mm}
\end{figure}

@@

Instead of multitasking multiple separate kernels, we merge a set of
kernels into a megakernel---a single, monolithic kernel.  The
megakernel is launched with as many workgroups as can be occupant
concurrently.  One workgroup takes the role of scheduler, and the
scheduling logic is embedded as part of the megakernel.  The remaining
workgroups act as a pool of workers.  A worker repeatedly queries the
scheduler to be assigned a task.  A task corresponds to executing a
cooperative or non-cooperative kernel.  In the non-cooperative case,
the workgroup executes the relevant kernel function uninterrupted,
then awaits further work.  In the cooperative case, the workgroup
either starts from the kernel entry point or immediately jumps to a
designated point within the kernel, depending on whether the workgroup
is an initial workgroup of the kernel, or a forked workgroup.  In the
latter case, the new workgroup also receives a struct containing the
values of all relevant $\transmit$-annotated variables.

@@

@@

The assignment mutex additionally protects concurrent calls to
$\offerfork$ and $\offerkill$.  On calling $\offerkill$, a workgroup
is only permitted to exit if the scheduler workgroup currently holds
the assignment mutex; that is, the scheduler is trying to obtain work
groups to send to a task.  Conversely, on calling $\offerfork$ a
workgroup can only spawn new workgroups if the scheduler workgroup
does not hold the assignment mutex.  Thus the assignment mutex, only
ever acquired by the scheduling workgroup, \emph{indirectly} provides
mutual exclusion between a pairs of workgroups executing $\offerkill$
and $\offerfork$ concurrently.

@@

\begin{table}[t]
\normalsize
\centering
\begin{tabular}{ l r r r r r r}
App. & barriers & kill & fork & transmit & LoC & inputs\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
%\rowcolor{Gray1}
%bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
%\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\vspace{-10mm}
\end{table}

@@

Our prototype implementation (\mysec\ref{sec:implementation}) requires
two optional features of OpenCL 2.0: SVM fine-grained buffers and SVM
atomics. This rules out evaluation on \nvidia and ARM Mali GPUs at
present, since these vendors do not yet support OpenCL $2.0$,
equivalent functionality is not provided by \nvidia's CUDA. Intel
drivers meet our requirements on Windows, but not on Linux.
%(we
%confirmed this by trying the upstream driver that requires a patched
%Linux kernel).
AMD support SVM atomics only their Kaveri family of
chips and only on Linux.%, requiring to compile the latest driver.

@@

\begin{figure*}
\includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf}
\includegraphics[width=.67\columnwidth]{iris_bfs_usa.pdf}
\includegraphics[width=.67\columnwidth]{iris_color_G3_circuit.pdf}
%\vspace{-2mm}
\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
\vspace{-3mm}
\end{figure*}

\begin{table}
\normalsize
\centering
\begin{tabular}{ l l | l l | l l }
\multicolumn{2}{c|}{overall} & \multicolumn{2}{c|}{barrier} & \multicolumn{2}{c}{wk.steal.} \\
mean & max & mean & max & mean & max \\
\hline
$1.07$ & $1.23^{\ddagger}$ & $1.06$ & $1.20^{\diamond}$ & $1.12$ & $1.23^{\ddagger}$ \\
\end{tabular}\\
{\small
 $^{\ddagger}$octree, $^{\diamond}$color G3\_circuit
}
%\vspace{-2mm}
\caption{Cooperative kernel slowdown w/o multitasking}
\label{tab:overhead}
\vspace{-8mm}
\end{table}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{heavy.pdf}
%\vspace{-2mm}
\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
\vspace{-5mm}
\end{figure}

>>>>

\newcommand{\ADComment}[1]{\textcolor{green}{AD: #1}}
%\newcommand{\TSComment}[1]{\textcolor{blue}{TS: #1}}
\newcommand{\TSAdded}[1]{\textcolor{blue}{#1}}
\newcommand{\TSEmph}[1]{\textcolor{red}{#1}}
\newcommand{\HEComment}[1]{\textcolor{orange}{HE: #1}}
%\newcommand{\COULDCUT}[1]{\textcolor{blue}{#1}}
\newcommand{\TSComment}[1]{\textcolor{purple}{TS: #1}}

@@

@@

% lstlisting macros
\lstset{
  language=C, % choose the language of the code
  showspaces=false, % show spaces adding particular underscores
  showstringspaces=false, % underline spaces within strings
  showtabs=false, % show tabs within strings adding particular underscores
  tabsize=8, % sets default tabsize to 2 spaces
  captionpos=b, % sets the caption-position to bottom
  mathescape=true, % activates special behaviour of the dollar sign
  basicstyle=\footnotesize\tt,
  columns=fullfelxible,
  xleftmargin=2em,
  %commentstyle=\rmfamily\itshape,
  morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then,transmit},
  escapeinside={(*@}{@*)},
  numbers=left
}

@@

\author{Tyler Sorensen}
\affiliation{%
  \institution{Imperial College London}
  \city{London}
  \country{UK}}
\email{t.sorensen15@imperial.ac.uk}

\author{Hugues Evrard}
\affiliation{%
  \institution{Imperial College London}
  \city{London}
  \country{UK}}
\email{h.evrard@imperial.ac.uk}

\author{Alastair F. Donaldson}
\affiliation{%
  \institution{Imperial College London}
  \city{London}
  \country{UK}}
\email{alastair.donaldson@imperial.ac.uk}

@@

%\authorinfo{Tyler Sorensen} {\makebox{Imperial College London, UK} } {t.sorensen15@imperial.ac.uk }

@@

\newcommand{\myparagraph}[1]{\paragraph{#1.}}
%\newcommand{\myparagraph}[1]{\vspace{1mm}\noindent\emph{\textbf{#1.}}}
%{\noindent{\bf\emph{#1}}\hspace{1em}}

@@

\myparagraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many interesting data-parallel algorithms are \emph{irregular}: the
amount of work to be processed is unknown ahead of time and may be
determined by the computation itself.  There is growing interest in
accelerating such algorithms on
GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}. \TSAdded{Such
  algorithms include octree partitioning, which has applications in
  particle simulations; and graph traversals, which have applications
  in speech decoding.}

@@

%\vspace{-1mm}
\myparagraph{A degree of fairness: occupancy-bound execution} The current GPU programming
models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
HSA~\cite{HSAprogramming11}, specify almost no guarantees regarding
scheduling of workgroups, and current GPU schedulers are unfair in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.\footnote{In practice,
  depending on the kernel, multiple workgroups might map to the same
  compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle more workgroups being
launched than there are compute units is via an \emph{occupancy-bound}
execution
model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16} where,
once a workgroup has commenced execution on a compute unit (it has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it finishes execution.
This simple model is easy to implement
efficiently, and experiments suggest that
it is widely employed by today's
GPUs~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{overview.pdf}
\caption{Cooperative kernels can flexibly resize to let other tasks
(e.g. graphics) run concurrently}
\label{fig:overview}
%\vspace{-2mm}
\end{figure}

@@

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that GPU vendors (1) will
not commit to the occupancy-bound execution model they currently
implement, for the above reasons, yet (2) will not guarantee fair
scheduling \TSAdded{using preemption. This is due to the high runtime
  cost of preempting workgroups on GPUs, which requires managing
  thread local state (e.g. registers, program location) for all
  threads in a workgroup (up to 1024 on Nvidia GPUs), as well as
  \emph{shared memory}, the workgroup local cache (up to 64 KB on
  Nvidia GPUs)}.  Vendors instead wish to retain the essence of the
  simple occupancy-bound model, supporting preemption only in key
  special cases.

@@

A kernel
that requires fair scheduling is identified as \emph{cooperative}, and written using two additional
language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.
%
Where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates that the workgroup
would ideally continue executing, but that the
scheduler may preempt the workgroup; the cooperative kernel
must be prepared to deal with either scenario.
%
Where the cooperative kernel could use additional resources, a workgroup can execute
$\offerfork$ to indicate that the
kernel is prepared to proceed with the existing set of workgroups, but
is able to benefit from one or more additional workgroups
commencing execution directly after the $\offerfork$ program point.

@@

\TSAdded{Cooperative kernels allow for \emph{cooperative multitasking}
  (see \mysec{\ref{sec:relatedwork}}), a multitasking approach used
  historically when preemption was not available or, in the case of
  GPUs, too costly (see previous paragraph). Our approach avoids the
  cost of preemptive multitasking as the state of a workgroup killed
  via $\offerkill$ does not have to be saved. Previous cooperative
  multitasking systems often provide \emph{yield} based semantics,
  where a processing unit would temporarily give up its hardware
  resource. We intentionally deviate from this design as in the case
  of a global barrier (a common idiom in the applications we study),
  adopting yield semantics would force the cooperative kernel to block
  \emph{completely} in the case where a single workgroup yields,
  stalling the kernel until the given workgroup resumes.  Instead, our
  $\offerkill$ allows a kernel to make progress with a smaller number
  of workgroups, with workgroups potentially joining again later via
  $\offerfork$.}

@@

In summary, our main contributions are:
\emph{cooperative kernels}, an extended GPU programming model that supports the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}); a \emph{prototype implementation} of cooperative
  kernels on top of OpenCL 2.0
  (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms \cutthree{across three GPUs} (\mysec\ref{sec:experiments}), including a best-effort comparison with the efficiency afforded by hardware-supported preemption available on Nvidia GPUs.

@@

%\ADComment{Reviewer comment: Section 2.2 would benefit from a short introduction explaining how the proposed examples fail in current implementations.}

@@

% lstlisting macros

\begin{lstlisting}
(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * queues) {
(*@\label{line:wksteal:getgroupid}@*)  int queue_id = get_group_id();
(*@\label{line:wksteal:mainloop}@*)  while (more_work(queues)) {
(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
    if (t)
(*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
%\vspace{-6mm}
\end{figure}

@@

\myfiglong\ref{fig:workstealing} illustrates a work stealing
kernel. Each thread receives a pointer to the task queues, in global
memory, initialized by the host to contain initial tasks. A thread
uses its workgroup id (line~\ref{line:wksteal:getgroupid}) as a queue
id to access the relevant task queue. The $\mathsf{pop\_or\_steal}$
function (line~\ref{line:wksteal:poporsteal}) pops a task from the
workgroup's queue or tries to steal a task from other queues. Although
not depicted here, concurrent accesses to queues inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex per queue, implemented using atomic compare and swap operations
on global memory.

If a task was obtained, then the workgroup processes it
(line~\ref{line:wksteal:processtask}), which may lead to new tasks
being created and pushed to the workgroup's queue. The kernel presents
two opportunities for spin-waiting: spinning to obtain a mutex, and
spinning in the main kernel loop to obtain a task.

%
% is called by all threads of a workgroup.  Inside
% $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
% the \emph{master thread}, tries to pop a
% task from the workgroup's queue, or to steal a task from a different
% queue. The master thread uses local memory inside
% $\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
% all threads of its workgroup.
%

@@

\begin{lstlisting}
kernel graph_app(global graph * g,
                 global nodes * n0, global nodes * n1) {
  int level = 0;
  global nodes * in_nodes = n0;
  global nodes * out_nodes = n1;
  int tid = get_global_id();
  int stride = get_global_size();
(*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
    for (int i = tid; i < in_nodes.size; i += stride)
      process_node(g, in_nodes[i], out_nodes, level);
(*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
(*@\label{line:graph:gb1}@*)    global_barrier();
(*@\label{line:graph:reset}@*)    reset(out_nodes);
    level++;
(*@\label{line:graph:gb2}@*)    global_barrier();
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{An OpenCL graph traversal algorithm}\label{fig:graphsearch}
%\vspace{-2mm}
\end{figure}

@@

The global barrier used in this application is not provided as a GPU
primitive, though previous works have shown that such a global barrier
can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16}, based
on CPU barrier designs~\cite[ch. 17]{HS08}.  These barriers employ
spinning to ensure threads wait at the barrier until all threads have
arrived, which requires fair scheduling between workgroups for the
barrier to operate correctly.

The mutexes and barriers used by these two examples appear to run
reliably on current GPUs for kernels that are executed with no more
workgroups than there are compute units.  This is due to the fairness
of the occupancy-bound execution model that current GPUs have been
shown, experimentally, to provide.  But, as discussed in
\mysec\ref{sec:intro}, this model is not endorsed by language
standards or vendor implementations, and is unlikely to be respected
in the future. \TSAdded{Without a fair schedule, the mutexes and
  barriers, which use spin-waiting in their implementation, are not
  guaranteed to be satisfied. For example, if a workgroup is not
  fairly scheduled after obtaining a mutex, there is no guarantee that
  the mutex will be released. Thus the kernel may hang while other
  workgroups wait indefinitely to obtain the mutex. }
%

In \mysec\ref{sec:programmingguidelines} we show how the work stealing
and graph traversal examples of \myfigs\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
updated to use our cooperative kernels programming model to resolve
the scheduling issue.

@@

%\vspace{-1mm}
\myparagraph{Active workgroups}
%
If the host application launches a cooperative kernel requesting $N$
workgroups, this indicates that the kernel should be executed with a
maximum of $N$ workgroups, and that as many workgroups as possible, up
to this limit, are desired.  However, the scheduler may initially
schedule fewer than $N$ workgroups, and as explained below the number
of workgroups that execute the cooperative kernel can change during
the lifetime of the kernel.

@@

\begin{lstlisting}
kernel graph_app(global graph *g,
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < in_nodes.size; i += stride)
      process_node(g, in_nodes[i], out_nodes, level);
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
%\vspace{-4mm}
\end{figure}

@@

Instead of multitasking multiple separate kernels, we merge a set of
kernels into a megakernel---a single, monolithic kernel.  The
megakernel is launched with as many workgroups as can be occupant
concurrently.  One workgroup takes the role of the scheduler, and the
scheduling logic is embedded as part of the megakernel.  The remaining
workgroups act as a pool of workers.  A worker repeatedly queries the
scheduler to be assigned a task.  A task corresponds to executing a
cooperative or non-cooperative kernel.  In the non-cooperative case,
the workgroup executes the relevant kernel function uninterrupted,
then awaits further work.  In the cooperative case, the workgroup
either starts from the kernel entry point or immediately jumps to a
designated point within the kernel, depending on whether the workgroup
is an initial workgroup of the kernel, or a forked workgroup.  In the
latter case, the new workgroup also receives a struct containing the
values of all relevant $\transmit$-annotated variables.

@@

\TSAdded{We note that the scheduler requirements given in
  \mysec{\ref{sec:cooperativekernels}} are agnostic to whether the
  scheduling logic takes place on the CPU or GPU. To
  avoid expensive communication between GPU and host, we choose to
  implement the scheduler on the GPU.}

@@

The assignment mutex additionally protects concurrent calls to
$\offerfork$ and $\offerkill$.  On calling $\offerkill$, a workgroup
is only permitted to exit if the scheduler workgroup currently holds
the assignment mutex; that is, the scheduler is trying to obtain work
groups to send to a task.  Conversely, on calling $\offerfork$ a
workgroup can only spawn new workgroups if the scheduler workgroup
does not hold the assignment mutex.  Thus the assignment mutex, only
ever acquired by the scheduling workgroup, \emph{indirectly} provides
mutual exclusion between pairs of workgroups executing $\offerkill$
and $\offerfork$ concurrently.

@@

\begin{table}[t]
\normalsize
\centering
\begin{tabular}{ l r r r r r r}
App. & barriers & kill & fork & transmit & LoC & inputs\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
%\rowcolor{Gray1}
%bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
%\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
%\vspace{-10mm}
\end{table}

@@

Our prototype implementation (\mysec\ref{sec:implementation}) requires
two optional features of OpenCL 2.0: SVM fine-grained buffers and SVM
atomics. This rules out evaluation on \nvidia and ARM Mali GPUs at
present, since these vendors do not yet support OpenCL $2.0$,
equivalent functionality is not provided by \nvidia's CUDA. Intel
drivers meet our requirements on Windows, but not on Linux.
%(we
%confirmed this by trying the upstream driver that requires a patched
%Linux kernel).
AMD supports SVM atomics only in their Kaveri family of
chips and only on Linux.%, requiring to compile the latest driver.

@@

\begin{figure*}
\includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf}
\includegraphics[width=.67\columnwidth]{iris_bfs_usa.pdf}
\includegraphics[width=.67\columnwidth]{iris_color_G3_circuit.pdf}
%\vspace{-2mm}
\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
%\vspace{-3mm}
\end{figure*}

\begin{table}
\normalsize
\centering
\begin{tabular}{ l l | l l | l l }
\multicolumn{2}{c|}{overall} & \multicolumn{2}{c|}{barrier} & \multicolumn{2}{c}{wk.steal.} \\
mean & max & mean & max & mean & max \\
\hline
$1.07$ & $1.23^{\ddagger}$ & $1.06$ & $1.20^{\diamond}$ & $1.12$ & $1.23^{\ddagger}$ \\
\end{tabular}\\
{\small
 $^{\ddagger}$octree, $^{\diamond}$color G3\_circuit
}
%\vspace{-2mm}
\caption{Cooperative kernel slowdown w/o multitasking}
\label{tab:overhead}
%\vspace{-8mm}
\end{table}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{heavy.pdf}
%\vspace{-2mm}
\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
%\vspace{-5mm}
\end{figure}
